---
title: "02 Kaplan-Meier Analysis"
author: "Kevin Schoelz"
date: "August 2019"
output:
  tufte::tufte_html:
    toc: true
    toc_depth: 2
---

We have an experiment where we are looking at the behavior of different ecotypes of a certain species of plant. In the experiment, the plant are inoculated by a virus. They are then observed once a day for several days to see when virus symptoms become apparent. There is reason to believe that maybe some of the ecotypes can slow the infection of the plant. In this notebook, we want to analyze the data using the Kaplan Meier method.

```{r setup, echo=FALSE, message=FALSE}

# Clear the environment
rm(list = ls())
#gc(reset = TRUE, verbose = FALSE)

# Notebook Settings
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(broom)
library(tufte)
library(cowplot)

#ggplot2 settings
theme_set(theme_cowplot())

```

We can start by loading up the data. There are three datafiles that we need to import from the previous notebook

1. `tidyplants.csv`: Rows correspond to data on individual plants, columns contain variables
2. `tidyplants_cdf.csv`: File has been rewritten in terms of what observations happened on a given day

In the grid of plots below, we are looking at the number of new infections observed on a given day. The data has been divided up so that different columns of plots show the behavior of a given ecotype. The control plants are in the left-most column. The different rows correspond to different trials. Each trial consists of plants that were inoculated at the same time. We expect that the different trials are probably not directly comparable due to uncontrolled environmental effects.

```{r load-data, echo=FALSE}

# Load data about new.infections pdf from csv file
raw_df <- read.csv("tidyplants.csv")
raw_cdf <- read.csv("tidyplants_cdf.csv")

head(raw_df)

head(raw_cdf)

```

# Kaplan Meier Analysis

## Survival Models

In a survival analysis, we have some object, (people, plants, machine etc...) that can undergo some type of event. In a survival analysis, our goal is to estimate the time until the event happens. The canonical example would be to analyze mortality curves, i.e. to estimate time until death for people who undergo some type of treatment. But the mathematical framework is more general than that. We can use survival analysis to analyze time until a disease shows up (as we will do here), or even time until a given part fails on a machine. It is a very general and powerful framework. 
In survival analysis, we are trying to estimate a series of functions. We have the probability distribution function $f(y)$ which gives the probability of a new infection occuring on day $y$. There is the cumulative distribution function, $F(y)$ which gives the probability that the event has occured by day $y$. Finally we have the survival function $S(y) = 1 - F(y)$ which gives the probability of a given individual surviving until day $y$. Classically, the goal of a survival analysis is to estimate the survival function $S(y)$.

## Estimating the survival function using the Kaplan-Meier method

How do we estimate the survival function? This is a big question, and there are a lot of different approaches. Let's generate some fake survival data to try and get a better feel for how to answer this question. Suppose we have some time until event data. This data is drawn from the following true survival function, as given below.

We want to estimate the survival function from our data. We can use the `rweibull` function to draw samples from a Weibull distribution. Results are stored in the table below.

```{r simulate-data, echo=FALSE}

y <- seq(1, 15, 0.1)

alpha = 10
sigma = 10



S <- exp(-(y/sigma)^alpha)


surv_ex_df <- data.frame("Time" = y, "Survival" = S)

# Simulate data from Weibull distribution
n_samples <- 20               # 20 individuals
set.seed(1234)                # set seed for reproducibility
y_samples <- floor(rweibull(n_samples, shape = alpha, scale = sigma))

# Store in data frame
simulate_df <- data.frame(ysim = y_samples) %>%
  arrange(ysim) %>%
  mutate(n_infected = row_number(),
         n_plants = 20,
         n_safe = n_plants - n_infected)

# Transform from new.infections to survival
surv_df <- simulate_df %>%
  group_by(ysim) %>%
  tally(name = "new_infections") %>%
  mutate(n_plants = 20,
         n_infected_total = cumsum(new_infections),
         n_healthy = n_plants + new_infections - n_infected_total,
         surv_prob = 1 - new_infections/n_healthy,
         cum_surv_prob = cumprod(surv_prob),
         proportion = 1 - n_infected_total/n_plants)


knitr::kable(surv_df)

```

```{r survival_figure_example, echo=FALSE, fig.margin = TRUE, fig.cap="Solid line indicates the Kaplan-Meier estimate of the survival curve. Dashed line indicates 'true' survival curve."}

ggplot(data = surv_df) +
  geom_step(aes(x = ysim, y = cum_surv_prob), color = "DodgerBlue4", size = 1) +
  geom_line(data = surv_ex_df, 
            aes(x = Time, y = surv_ex_df$Survival), 
            color = "DodgerBlue4", size = 1, 
            linetype = "dashed") + 
  xlim(6,14) +
  xlab("Time (Days)") + 
  ylab("Survival Probability")

```


A plot of the Survival function is shown in the margin. The dashed line shows the "true" survival function, while the solid line shows the Kaplan-Meier estimate as generated from the observed data. Note, for the Kaplan-Meier estimate, we use a stepping function, as there is no empirical reason to change the probability until we see failure events.

## Comparing survival curves using the log-rank test

Suppose we are taking survival data from two different groups. This might be a treatment and a control group. We would like to be able to say whether or not the groups seem distinct. A non-parametric approach would be to use the Log-Rank test. In this test, at each time step we look at new infections in each group compared to the total healthy population in the time before. We also look at the two groups together. In this way we can see how the statistics of the two groups compare to the statistics of a single group containing all of the individuals. We can estimate the probability that one group could produce these results using the $\chi^{2}$ statistic.

$$
\chi^{2} = \frac{\left(E_{A} - O_{A}\right)^{2}}{E_{A}} + \frac{\left(E_{B} - O_{B}\right)^{2}}{E_{B}}
$$

For a single timestep like this, we have one degree of freedom. We can use the $\chi^{2}$ test statistic in order to compute the liklihood that both groups come from the same distribution. We can calculate a $p$ value by comparing the two groups, and reject the null hypothesis (that the two groups are equivalent) if $p < 0.05$.

# Kaplan Meier Analysis

With some slight changes to the `tidyplants_cdf.csv` file we can get the Kaplan Meier analysis.

```{r}

tidyplants_cdf <- read.csv("tidyplants_cdf.csv")

tidyplants_cdf <- tidyplants_cdf %>%
  group_by(trial_name, ecotype) %>%
  mutate(new_infections = ifelse(day == 1, 0, total_infections - lag(total_infections)),
         healthy_pool = ifelse(day == 1, Ntot, Ntot - lag(total_infections)),
         prop = 1 - new_infections/healthy_pool,
         km_survival = cumprod(prop),
         km_helpercol1 = new_infections/(healthy_pool + new_infections),
         km_stan_error = km_survival*sqrt(cumsum(km_helpercol1)),
         km_95conf = 1.96*km_stan_error,
         km_95upper = km_survival + km_95conf,
         km_95lower = km_survival - km_95conf)

treatments <- tidyplants_cdf %>%
  filter(ecotype != "Col") %>%
  rename(total_infections_treat = total_infections,
         Ntot_treat = Ntot,
         Nobs_treat = Nobs,
         Ncens_treat = Ncens,
         ycens_treat = ycens,
         new_infections_treat = new_infections,
         healthy_pool_treat = healthy_pool,
         prop_treat = prop,
         km_survival_treat = km_survival,
         km_helpercol1_treat = km_helpercol1,
         km_stan_error_treat = km_stan_error,
         km_95conf_treat = km_95conf,
         km_95upper_treat = km_95upper,
         km_95lower_treat = km_95lower)

controls <- tidyplants_cdf %>%
  filter(ecotype == "Col") %>%
  rename(total_infections_ctrl = total_infections,
         Ntot_ctrl = Ntot,
         Nobs_ctrl = Nobs,
         Ncens_ctrl = Ncens,
         ycens_ctrl = ycens,
         new_infections_ctrl = new_infections,
         healthy_pool_ctrl = healthy_pool,
         prop_ctrl = prop,
         km_survival_ctrl = km_survival,
         km_helpercol1_ctrl = km_helpercol1,
         km_stan_error_ctrl = km_stan_error,
         km_95conf_ctrl = km_95conf,
         km_95upper_ctrl = km_95upper,
         km_95lower_ctrl = km_95lower)

km_cdf <- left_join(treatments, controls, by = c("trial_name", "day"))

km_cdf <- km_cdf %>%
  mutate(Ntot_total = Ntot_ctrl + Ntot_treat,
         new_infections_total = new_infections_ctrl + new_infections_treat,
         healthy_pool_total = healthy_pool_treat + healthy_pool_ctrl)


km_plot <- ggplot(data = km_cdf) + 
  geom_step(aes(x = day, y = km_survival_treat), color = "Maroon", size = 1) +
  geom_step(aes(x = day, y = km_survival_ctrl), color = "DodgerBlue4", size = 1) +
  facet_grid(trial_name ~ ecotype.x) +
  xlim(6,13) +
  ylab("Survival Probability") +
  xlab("Day")

km_plot

```


# Using the `survival` and `survminer` packages to estimate the survival function

Below is the code to calculate the survival function for the survival packages.

```{r}

tidyplants <- read.csv("tidyplants.csv")

trial_names_list <- list("Trial1LL",
                         "Trial2LL",
                         "Trial3LL",
                         "Trial4LL")

ecotypes_list <- list("VIII-A",
                      "VIII-B",
                      "XI-2",
                      "XI-A",
                      "XI-B",
                      "XI-K")

for (eco in ecotypes_list){
  for (trial in trial_names_list){
    temp_df <- tidyplants %>% filter((ecotype == eco | ecotype == "Col"), trial_name == trial)
    surv_obj <- Surv(time = temp_df$first_infection, event = ifelse(temp_df$censored == 0, 1, 0))
    fit <- survfit(formula = surv_obj ~ temp_df$ecotype, data = temp_df)
  }
}

```









